Algorithm explanation

### 2013-05-22 article/algo.tex
Disjunction algorithm construction

The Horn clause solving algorithm receives an input problem in Horn
graph. A \textbf{Horn graph} is a labelled directed graph $G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two disjoint sets of vertices; Horn term
  vertices $V_T$ and disjunction vertices $V_\vee$. Each vertex $u \in
  V_\vee$ has eactly one outgoing edge to some $v \in V_T$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, \ldots, x_{arity(P)})$ or a linear inequality
  $a_1 x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}
There are two kinds of edges in a Horn graph;
\begin{itemize}
\item for $u \in V_T, v \in V_\vee$, an edge takes a form
  $(u,v,\theta)$, written $\edgel{u}{\theta}{v}$, or,
\item for $u \in V_\vee, v \in V_T$, an edge takes a form
  $(u,v,\emptyset)$, written $\edge{u}{v}$, where the mapping is
  empty.
\end{itemize}

@@

The solution of $G$ is a map $\rho$ from predicate variables to linear
predicates of the form $\lambda x_1, \cdots ,x_n. \psi $ (where $\psi$
is a formula constructed from
\begin{align*}
\psi ::= & a_1 x_1 + \cdots + a_n x_n \leq b \mid \\
& \psi \wedge \psi \mid \psi \vee \psi \mid \top \mid \bot
\end{align*}
), where $\rho[G]$ is a tautology. A solution is \textit{simple} if
$\rho[P]$ is of the form $a_1 x_1 + \cdots + a_n x_n \leq b$ for every
$P$ in $dom(\rho)$.

@@

For simplicity of discussion, we define \textbf{term arity} $k_v$
for every Horn term vertex $v \in V_T$.
\begin{align*}
k_v =
\begin{cases}
arity(P) & \mbox{if } \varphi(v) = P(x_1,...,x_{arity(P)}) \\
0 & \mbox{if } \varphi(v) = \top \mbox { or } \varphi(v) = \bot
\end{cases}
\end{align*}
We let $K$ denote the maximum of $k_v$ over all $v \in V_T$.

@@

\subsection{DAG problem}

@@

We give a constraint function over a copy tree
$\hat \Delta(v,e)$ for $v \in V_C, e \in E$ as
\[ \hat \Delta (v,e) = \textsc{Copy}_v (\Delta_{V'} (\xi (v)))
\text{ such that } V' = \left\lbrace v' \middle| (v, v',e) \in E_S \right\rbrace \]
extending $\Delta$ function to
$\Delta_V (v) = (\mathbf{a}_{v,\emptyset} \mathbf{x}_v \leq b_{v,\emptyset}, C_v)$
as
\begin{align*}
C_v = & \bigcup_\circ C_u \cup \bigcup_\bullet C_u \cup
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u \leadsto v \\
\left\lbrace
 \mathbf{a}_{v,\emptyset,i} =
  \sum_\circ \mathbf{a}_{u,U,\theta^{-1}(i)} +
  \sum_\bullet \mathbf{a}_{u,U, \theta^{-1}(i)}
\right\rbrace \cup \\
\hspace{2cm} \left\lbrace
 b_{v,\emptyset} \geq
  \sum_\circ b_{u,U} +
  \sum_\bullet b_{u,U}
\right\rbrace \cup \\
\hspace{2cm} \bigcup_{k_v < i \leq K} \left\lbrace \mathbf{a}_{v,\emptyset,i} = 0 \right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}_v) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_{v,\emptyset} = \lambda_v \mathbf{a}_0,
 b_{v,\emptyset} = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x}_v \leq b_0
\end{cases}
\end{align*}
under
\[
\circ = \substack{
  u \mathop{\leadsto}^\theta v \\
  \nexists u' \in V; \xi (u') = u \\
  \Delta_V (u) = (\mathbf{a}_{u,U} \mathbf{x}_u \leq b_{u,U}, C_u)}
\qquad \bullet = \substack{
  u \mathop{\leadsto}^\theta v \\
  \exists u' \in V; \xi (u') = u \\
  \hat \Delta (u', \edge{u}{v}) = (\mathbf{a}_{u,U} \mathbf{x}_u \leq b_{u,U}, C_u)} \]

@@

\subsection{Disjunctive problems}

We eliminate the restriction that all disjunction vertices $V_\vee$ have
exactly one predecessor. We define the universal set of selectors for
disjunction vertices.
\[ \mathcal{S} = \prod_{v \in V_\vee} \left\lbrace v \mapsto u \middle| \edgel{u}{\emptyset}{v} \right\rbrace \]

>>>>

The Horn clause solving algorithm receives an input problem in Horn
graph. A \textbf{Horn graph} is a labelled directed graph $G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two disjoint sets of vertices; Horn term
  vertices $V_T$ and disjunction vertices $V_\vee$. Each vertex $u \in
  V_\vee$ has eactly one outgoing edge to some $v \in V_T$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, \ldots, x_{\mathrm{arity}(P)})$ or a linear inequality
  $a_1 x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}
There are two kinds of edges in a Horn graph;
\begin{itemize}
\item for $u \in V_T, v \in V_\vee$, an edge takes a form
  $(u,v,\theta)$, written $\edgel{u}{\theta}{v}$, or,
\item for $u \in V_\vee, v \in V_T$, an edge takes a form
  $(u,v,\emptyset)$, written $\edge{u}{v}$, where the mapping is
  empty.
\end{itemize}

@@

The solution of $G$ is a map $\rho$ from predicate variables to linear
predicates of the form $\lambda x_1, \cdots ,x_n. \psi $ (where $\psi$
is a formula constructed from
\begin{align*}
\psi ::= & a_1 x_1 + \cdots + a_n x_n \leq b \mid \\
& \psi \wedge \psi \mid \psi \vee \psi \mid \top \mid \bot
\end{align*}
), where $\rho[G]$ is a tautology. A solution is \textit{simple} if
$\rho[P]$ is of the form $a_1 x_1 + \cdots + a_n x_n \leq b$ for every
$P$ in $\mathrm{dom}(\rho)$.

@@

For simplicity of discussion, we define \textbf{term arity} $k_v$
for every Horn term vertex $v \in V_T$.
\begin{align*}
k_v =
\begin{cases}
\mathrm{arity}(P) & \mbox{if } \varphi(v) = P(x_1,...,x_{\mathrm{arity}(P)}) \\
0 & \mbox{if } \varphi(v) = \top \mbox { or } \varphi(v) = \bot
\end{cases}
\end{align*}
We let $K$ denote the maximum of $k_v$ over all $v \in V_T$.

@@

\subsection{Disjunction-free problems}

@@

We give a constraint function over a copy tree
$\hat \Delta(v,e)$ for $v \in V_C, e \in E$ as
\[ \hat \Delta (v,e) = \textsc{Copy}_v (\Delta_{V'} (\xi (v)))
\text{ such that } V' = \left\lbrace v' \middle| (v, v',e) \in E_S \right\rbrace \]
extending $\Delta$ function to
$\Delta_{V'} (v) = (\mathbf{a}_{v,\emptyset} \mathbf{x}_v \leq b_{v,\emptyset}, C_v)$
as
\begin{align*}
C_v = & \bigcup_\circ C_u \cup \bigcup_\bullet C_u \cup
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u \leadsto v \\
\left\lbrace
 \mathbf{a}_{v,\emptyset,i} =
  \sum_\circ \mathbf{a}_{u,U,\theta^{-1}(i)} +
  \sum_\bullet \mathbf{a}_{u,U, \theta^{-1}(i)}
\right\rbrace \cup \\
\hspace{2cm} \left\lbrace
 b_{v,\emptyset} \geq
  \sum_\circ b_{u,U} +
  \sum_\bullet b_{u,U}
\right\rbrace \cup \\
\hspace{2cm} \bigcup_{k_v < i \leq K} \left\lbrace \mathbf{a}_{v,\emptyset,i} = 0 \right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}_v) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_{v,\emptyset} = \lambda_v \mathbf{a}_0,
 b_{v,\emptyset} = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x}_v \leq b_0
\end{cases}
\end{align*}
under the iterator $\circ = (u,U,\theta,C_u) \in S_v^- $ and $\bullet = (u,U,\theta,C_u) \in S_v^+ $
such that
\begin{align*}
& S_v^- = \left\lbrace (u, U, \theta, C_u) \middle|
  u \mathop{\leadsto}^\theta v \wedge
  \left( \nexists u' \in V'; \xi (u') = u \right) \wedge
  \Delta_{V'} (u) = (\mathbf{a}_{u,U} \mathbf{x}_u \leq b_{u,U}, C_u) \right\rbrace
\\
& S_v^+ = \left\lbrace (u, U, \theta, C_u) \middle|
  u \mathop{\leadsto}^\theta v \wedge
  \left( \exists u' \in V'; \xi (u') = u \wedge
  \hat \Delta (u', \edge{u}{v}) = (\mathbf{a}_{u,U} \mathbf{x}_u \leq b_{u,U}, C_u) \right) \right\rbrace
\\
\end{align*}

@@

\subsection{Acyclic Horn problems}

We eliminate the restriction that all disjunction vertices $V_\vee$ have
exactly one predecessor.

We define selectors for preceding disjunction vertices to $v$.
\[ \mathcal{S}_v = \prod_{u \in \mathrm{succ}(v)} \left\lbrace u \mapsto t \middle| \edgel{t}{\theta}{u} \right\rbrace \]

\begin{align*}
K_l = \begin{cases}
\left\lbrace l \right\rbrace & \mbox{if } l = \mathbf{a} \mathbf{x} \leq b \\
\mathcal{X}_P \cup \left\lbrace \bot \right\rbrace
& \mbox{if } l = P(\mathbf{x})
\end{cases} \\
\forall s \in \mathcal{S}_v;
X = \prod_{t \in \mathrm{ran}(s)} K_{\varphi(t)}
\exists c \in X;
\end{align*}
